The Power of Unbiased Recursive Partitioning:\\ A Unifying View of CTree, MOB, and GUIDE

### 2019-05-23 pkg/partykit/inst/guideliketest/draft/draft.Rnw
minor revisions, mainly in the appendix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2587 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 0d11ed92ff4d807656c2080c21605e0b2e9daee8

To test for independence of $h(Y)$ and $g(Z)$ CTree calculates a linear association test statistic,
following the framework of \cite{Strasser+Weber:1999}. The corresponding conditional expectation and
covariance given all permutations of the response variable can be calculated and used to 
standardize the test statistic. This standardized statistic has an asymptotic normal distribution
which is in fact multivariate if either of the transformation $h(Y)$ and/or $g(Z)$ is
multivariate. The actual test is carried out by mapping this standardized statistic to
the real line either by taking the absolute maximum or using a quadratic form -- with $p$-values
being computed from the analogous transformation of the normal distribution
\citep[see][for a hands-on introduction]{Hothorn+Hornik+VanDeWiel:2006}.

@@

For selecting a split variable MOB employs a score-based test that relies on the central limit
theorem for the parameter estimate $\hat{\beta}$. The test assesses whether the scores
-- when ordered by the potential split variable $Z$ -- fluctuate randomly around their
zero mean or differ systematically in certain subgroups. The latter would indicate a
parameter instability that could be captured by fitting separate models (optimizing $\ell$)
in the resulting subgroups. In case of a numeric split variable $Z$ both the score-based statistic
and the partitioned objective function $\ell$ are maximized over all possible splits in $Z$
(subject to certain minimal subgroup size constraints). Unlike the CTree framework, MOB
relies on classical unconditional inference. For more details see \cite{Zeileis+Hornik:2007}.

@@

To construct a statistical test two additional transformations are carried out:
(1)~The residuals $r(Y,X,\hat{\beta})$ are binarized at zero, yielding an indicator for positive vs.\ negative residuals.
(2)~Each possible split variable $Z$ is categorized, i.e., unless $Z$ is already categorical
it is split at its quartiles into four bins.
Subsequently, a $\chi^2$~test of independence is performed for the binarized residuals and
each categorized/categorical split variable.
After choosing the split variable showing the highest dependency by yielding the lowest 
$p$-value, the split point minimizing the overall goodness-of-fit measure is selected. 
Note that the split point in numeric variables $Z$ is searched over all possible splits,
not just the four bins that were constructed for the $\chi^2$~test.

@@

\begin{table}[h!]
\centering
\begin{tabular}{ l l l }
\hline
Building block               & Levels              & Transformation \\
\hline
Residuals vs.\ scores        & \textit{residuals}  & $r(Y,X,\hat{\beta})$\\
\quad ($h$ transformation)   & \textit{scores}     & $s(Y,X,\hat{\beta})$\\
\hline
Binarization                 & \textit{yes}        & $\mathds{1}_{[0,\infty)}(h)$\\
\quad (of $h$)               & \textit{no}         & $h$ without binarization\\
\hline
Categorization               & \textit{cat}        & $g(Z)=(0,1,0,0)$\\
\quad ($g$ transformation)   &                     & \quad indicating the assigned bin\\
                             & \textit{max}        & $g(Z)=(0,\ldots,0,1,\ldots,1)$\\
                             &                     & \quad indicating the crossed quantile\\
                             & \textit{lin}        & $g(Z)=Z$\\
\hline
\end{tabular}
\caption{Available levels of the building blocks.
\label{tab:levels}}
\end{table}

@@

\section*{Acknowledgments}
\hlc{An extended research stay of Torsten Hothorn in Innsbruck (August
2017 to January 2018) was financially supported by the Swiss National
Science Foundation, grant number SNF IZSEZ0 177091.}

@@

\section{Test statistics}
This section provides further information on the test statistics applied in CTree, MOB, 
and GUIDE. Most of the presented details have been extracted from the original papers 
\cite{Hothorn+Hornik+Zeileis:2006}, \cite{Zeileis+Hothorn+Hornik:2008}, 
and \cite{Loh:2002}, however, notation is adapted to the main manuscript
in order to allow for better comparison.

\subsection{CTree}
To measure the association of a response $Y$ and each possible split variable $Z$ the
CTree algorithm applies a linear test statistic $T$ which is excerpted from Section~3.1.
(``Variable selection and stopping criteria'') of the original paper \citep{Hothorn+Hornik+Zeileis:2006} and is of the following form:
$$
T(Y, Z, w) = \text{vec}\left( \sum_{i=1}^n w_i g(Z_{i})h(Y_i, (Y_1,\ldots,Y_n))^{\top}\right)
$$
where $w$ are optional weights, $g: \mathcal{Z} \rightarrow \mathbb{R}^{p}$ is a nonrandom transformation of the split variable $Z$ and the influence function $h: \mathcal{Y} \times \mathcal{Y}^n \rightarrow \mathbb{R}^q$ depends on the response $Y = (Y_1, \ldots, Y_n)$ in a permutation symmetric way. Moreover, by applying the ``vec'' operator the resulting $p~\times~q$ matrix is converted into a $p q$ column vector.
Following \cite{Strasser+Weber:1999}
the conditional expectation $\mu$ and covariance $\Sigma$ of a linear test statistic can 
be calculated and used to standardize an observed linear test statistic $t$ within a 
function $c$ mapping into the real line.
For example, the maximum of the absolute values of the standardized linear statistic
$$
c_{\max}(t,\mu,\Sigma) = \max_{k=1,\ldots,pq} \left| \frac{(t-\mu)}{\sqrt{(\Sigma)_{kk}}} \right|
$$
or a quadratic form
$$
c_{\text{quad}}(t, \mu, \Sigma) = (t - \mu)\Sigma^{+} (t - \mu)^{\top}
$$
can be considered where $\Sigma^{+}$ is the Moore-Penrose inverse of $\Sigma$.
Since the asymptotic conditional distribution of a linear test statistic $T$ is a 
multivariate normal with parameters $\mu$ and $\Sigma$ \citep{Strasser+Weber:1999}, 
the asymptotic distribution of $c_{\text{max}}$ is normal while the quadratic form 
$c_{\text{quad}}$ follows an asymptotic $\chi^2$~distribution. Based on this knowledge 
the corresponding $p$-values can be calculated easily.

\subsection{MOB}
The MOB algorithm employs an empirical fluctuation process $W$ to measure deviations of
the model scores $s$ from zero. As described in detail in Section~3.2. (``Testing for 
parameter instability'') of the original paper \citep{Zeileis+Hothorn+Hornik:2008}
this process is of the following form:
$$
W(t) = \hat{J}^{-1/2} n^{-1/2} \sum_{i=1}^{\lfloor nt \rfloor} s_{\sigma(Z_{i})} \qquad (0 \leq t \leq 1)
$$
where the model scores $s$ are sorted by the split variable Z by including the ordering permutation $\sigma(Z_{i})$. 
To scale this partial sum process an estimate $\hat{J}$ of the covariance matrix $\text{cov}(s(Y,X,\hat{\beta}))$ is included.
Following \cite{Zeileis+Hornik:2007} $W(t)$ converges to a Brownian bridge $W^0$ under the null
hypothesis of parameter stability. To obtain a test statistic a scalar functional $\lambda(\cdot)$ capturing the fluctuation in the empirical process can be applied and the corresponding asymptotic distribution of $\lambda(W(t))$ can be obtained by employing the same functional to the limit process, i.e., $\lambda(W^0)$. 
For more details, particularly on calculating the corresponding $p$-values see Section~3.2. of the original paper \citep{Zeileis+Hothorn+Hornik:2008}.

\subsection{GUIDE}
The test statistic of the $\chi^2$~test as applied in the GUIDE algorithm is
$$
X^2 = \frac{(O_{11} - E_{11})^2}{E_{11}} + \ldots + \frac{(O_{11} - E_{11})^2}{E_{11}}
$$
where $O_{ij}$ are the observed frequencies of observations in a certain combination of binarized residuals ($i=1,2$) and categorized split variables ($j = 1,\ldots,4$). $E_{ij}$ are the corresponding expected frequencies under independence $E_{ij} = \frac{O_{i \cdot} \cdot O_{\cdot j}}{n}$. Under the null hypothesis of independence the asymptotic distribution of $X^2$ is a $\chi^2$~distribution allowing for a straight-forward calculation of $p$-values.
If full model scores are used instead of residuals a sum of the test statistic over the number of distribution parameters is considered such that each 
summand corresponds to one column of the score matrix.

>>>>

To test for independence of $h(Y)$ and $g(Z)$ CTree calculates a linear association test statistic,
following the framework of \cite{Strasser+Weber:1999}. The corresponding conditional expectation and
covariance given all permutations of the response variable can be calculated and used to 
standardize the test statistic. This standardized statistic has an asymptotic normal distribution
which is in fact multivariate if either of the transformation $h(Y)$ and/or $g(Z)$ is
multivariate. The actual test is carried out by mapping this standardized statistic to
the real line either by taking the absolute maximum or using a quadratic form -- with $p$-values
being computed from the analogous transformation of the normal distribution
\citep[see][for a hands-on introduction and Appendix~\ref{app:teststat_ctree} for more details on the linear test statistic]{Hothorn+Hornik+VanDeWiel:2006}.

@@

For selecting a split variable MOB employs a score-based test that relies on the central limit
theorem for the parameter estimate $\hat{\beta}$. The test assesses whether the scores
-- when ordered by the potential split variable $Z$ -- fluctuate randomly around their
zero mean or differ systematically in certain subgroups. The latter would indicate a
parameter instability that could be captured by fitting separate models (optimizing $\ell$)
in the resulting subgroups. In case of a numeric split variable $Z$ both the score-based statistic
and the partitioned objective function $\ell$ are maximized over all possible splits in $Z$
(subject to certain minimal subgroup size constraints). Unlike the CTree framework, MOB
relies on classical unconditional inference. For more details see \cite{Zeileis+Hornik:2007} and 
Appendix~\ref{app:teststat_mob}.

@@

To construct a statistical test two additional transformations are carried out:
(1)~The residuals $r(Y,X,\hat{\beta})$ are binarized at zero, yielding an indicator for positive vs.\ negative residuals.
(2)~Each possible split variable $Z$ is categorized, i.e., unless $Z$ is already categorical
it is split at its quartiles into four bins.
Subsequently, a $\chi^2$~test of independence is performed for the binarized residuals and
each categorized/categorical split variable.
After choosing the split variable showing the highest dependency by yielding the lowest 
$p$-value, the split point minimizing the overall goodness-of-fit measure is selected. 
Note that the split point in numeric variables $Z$ is searched over all possible splits,
not just the four bins that were constructed for the $\chi^2$~test. More details on the
applied test statistic can be found in Appendix~\ref{app:teststat_guide}.

@@

\begin{table}[h!]
\centering
\begin{tabular}{ l l l }
\hline
Building block               & Levels              & Transformation \\
\hline
Residuals vs.\ scores        & \textit{residuals}  & $h(Y) = r(Y,X,\hat{\beta})$\\
\quad ($h$ transformation)   & \textit{scores}     & $h(Y) = s(Y,X,\hat{\beta})$\\
\hline
Binarization                 & \textit{yes}        & $\mathds{1}_{[0,\infty)}(h)$\\
\quad (of $h$)               & \textit{no}         & $h$ without binarization\\
\hline
Categorization               & \textit{cat}        & $g(Z)=(0,1,0,0)$\\
\quad ($g$ transformation)   &                     & \quad indicating the assigned bin\\
                             & \textit{max}        & $g(Z)=(0,\ldots,0,1,\ldots,1)$\\
                             &                     & \quad indicating the crossed quantile\\
                             & \textit{lin}        & $g(Z)=Z$\\
\hline
\end{tabular}
\caption{Available levels of the building blocks.
\label{tab:levels}}
\end{table}

@@

\section*{Acknowledgments}
An extended research stay of Torsten Hothorn in Innsbruck (August
2017 to January 2018) was financially supported by the Swiss National
Science Foundation, grant number SNF IZSEZ0 177091.

@@

\section{Test statistics}\label{app:teststat}
This section provides further information on the test statistics applied in CTree, MOB, 
and GUIDE. Most of the presented details have been extracted from the original papers 
\cite{Hothorn+Hornik+Zeileis:2006}, \cite{Zeileis+Hothorn+Hornik:2008}, 
and \cite{Loh:2002}, however, notation is adapted to the main manuscript
in order to allow for better comparison.

\subsection{CTree}\label{app:teststat_ctree}
To measure the association of a response $Y$ and each possible split variable $Z$ the
CTree algorithm applies a linear test statistic $T$ which is excerpted from Section~3.1.
(``Variable selection and stopping criteria'') of the original paper \citep{Hothorn+Hornik+Zeileis:2006} and is of the following form:
$$
T(Y, Z, w) = \text{vec}\left( \sum_{i=1}^n w_i g(Z_{i})h(Y_i)^{\top}\right)
$$
where $w$ are optional weights, $g: \mathcal{Z} \rightarrow \mathbb{R}^{p}$ is a nonrandom 
transformation of the split variable $Z$ and the influence function 
$h: \mathcal{Y} \times \mathcal{Y}^n \rightarrow \mathbb{R}^q$ depends on the response 
$Y = (Y_1, \ldots, Y_n)$ in a permutation symmetric way and is set to 
$h(Y_i) = s(Y_i, X_i, \hat{\beta})$ for a score-based approach. 
Moreover, by applying the ``vec'' operator the resulting
$p~\times~q$ matrix is converted into a $p q$ column vector.
Following \cite{Strasser+Weber:1999}
the conditional expectation $\mu$ and covariance $\Sigma$ of a linear test statistic can 
be calculated and used to standardize an observed linear test statistic $t$ within a 
function $c$ mapping into the real line.
For example, the maximum of the absolute values of the standardized linear statistic
$$
c_{\max}(t,\mu,\Sigma) = \max_{k=1,\ldots,pq} \left| \frac{(t-\mu)}{\sqrt{(\Sigma)_{kk}}} \right|
$$
or a quadratic form
$$
c_{\text{quad}}(t, \mu, \Sigma) = (t - \mu)\Sigma^{+} (t - \mu)^{\top}
$$
can be considered where $\Sigma^{+}$ is the Moore-Penrose inverse of $\Sigma$.
Since the asymptotic conditional distribution of a linear test statistic $T$ is a 
multivariate normal with parameters $\mu$ and $\Sigma$ \citep{Strasser+Weber:1999}, 
the asymptotic distribution of $c_{\text{max}}$ is normal while the quadratic form 
$c_{\text{quad}}$ follows an asymptotic $\chi^2$~distribution. Based on this knowledge 
the corresponding $p$-values can be calculated easily.

\subsection{MOB}\label{app:teststat_mob}
The MOB algorithm employs an empirical fluctuation process $W$ to measure deviations of
the model scores $s$ from zero. As described in detail in Section~3.2. (``Testing for 
parameter instability'') of the original paper \citep{Zeileis+Hothorn+Hornik:2008}
this process is of the following form:
$$
W(t) = \hat{J}^{-1/2} n^{-1/2} \sum_{i=1}^{\lfloor nt \rfloor} s_{\sigma(Z_{i})} \qquad (0 \leq t \leq 1)
$$
with the model scores $s_i = s(Y_i, X_i, \hat{\beta}$ being sorted by the split variable Z by including the ordering permutation $\sigma(Z_{i})$. 
To scale this partial sum process an estimate $\hat{J}$ of the covariance matrix $\text{cov}(s(Y,X,\hat{\beta}))$ is included.
Following \cite{Zeileis+Hornik:2007} $W(t)$ converges to a Brownian bridge $W^0$ under the null
hypothesis of parameter stability. To obtain a test statistic a scalar functional $\lambda(\cdot)$ capturing the fluctuation in the empirical process can be applied and the corresponding asymptotic distribution of $\lambda(W(t))$ can be obtained by employing the same functional to the limit process, i.e., $\lambda(W^0)$.

One possible and intuitive choice for a functional in order to asses instabilities over a numerical 
split variable is the following:
$$\lambda_{supLM}(W) = \max_{i = \underline{i},\ldots,\overline{i}}\left(\frac{i}{n}\cdot \frac{n-i}{n}\right)^{-1} \left|\left|W\left(\frac{i}{n}\right)\right|\right|_2^2
$$
where a minimal segment size $\underline{i}$ and then $\overline{i} = n - \underline{i}$ are used to 
define the interval $[\underline{i},\overline{i}]$. Other possible functionals, for example also
for categorical split variables, and more details,
particularly on calculating the corresponding $p$-values can be found in Section~3.2. of the 
original paper \citep{Zeileis+Hothorn+Hornik:2008}.

\subsection{GUIDE}\label{app:teststat_guide}
The test statistic of the $\chi^2$~test as applied in the GUIDE algorithm is
$$
X^2 = \frac{(O_{11} - E_{11})^2}{E_{11}} + \ldots + \frac{(O_{24} - E_{24})^2}{E_{24}}
$$
where $O_{kl}$ are the observed frequencies of observations in a certain combination of binarized residuals ($k=1,2$) and categorized split variables ($l = 1,\ldots,4$). $E_{kl}$ are the corresponding expected frequencies under independence $E_{kl} = \frac{O_{k \cdot} \cdot O_{\cdot l}}{n}$. Under the null hypothesis of independence the asymptotic distribution of $X^2$ is a $\chi^2$~distribution allowing for a straight-forward calculation of $p$-values.
If full model scores are used instead of residuals a sum of the test statistic over the number of distribution parameters is considered such that each 
summand corresponds to one column of the score matrix.
